Evaluating Recommender Behavior For New Users

### 2014-07-23 paper/main.tex
another proofreading and formatting pass

  The ideal way to study this issue would be a live study with real people who are new to a recommender system.
  Users could be randomly assigned an algorithm and then surveyed after a random number of ratings to establish what they thought of the new user experience. 
  Unfortunately, there are many algorithms and new users are a scarce resource.

  Fortunately, there has been a growing trend in recommender system research to perform multiple metric offline evaluations of recommender systems.
  This work comes out of a pushback against early recommender evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
  The response of this pushback has been the generating of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
  %TODO: I want to cite more papers here, I don't think I have good ones either. Joe would probably know.
  This strategy allows us to dig much deeper into the differences between algorithms, looking past simple measures of accuracy to more meaningful dimensions such as how novel the recommendations are, or how frequently the recommendations change for any user.
  This allows us to capture the qualitative tradeoffs between algorithms which are otherwise quite similar, which helps us choose algorithms that are better suited for a particular domain.

@@

  Ideally we would measure new user experience with a live survey on users who are just joining a recommender service.
  Because new users are a scarce resource we opt to first explore the new user experience in an offline setting.
  This allows us to test many more profile sizes and algorithms much faster than an online study.
  We will adapt the standard train and test methodology so that test users have only a specified number of training ratings.
  By running this approach for various profile sizes we can generate plots of performance against profile size.
  This lets us understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.

@@

  Due to the low profile sizes involved in our evaluation, not every algorithm was able to make predictions for every user-item pair.
  In these cases, we use the UserItemBaseline prediction as a fallback.
  This ensured that our prediction based metrics were always evaluated over the same number of items.
  We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
  Therefore we form recommendations as the top 20 items that the algorithm was able to predict which were not in the training set.
  More information about these algorithms can be found in table \ref{tbl:algo}.
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

@@

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using the Ekstrand's values as a starting point.
  We tuned based on RMSE performance.
  While tuning, we found that the previous damping parameter used for the baselines to be too large.
  We found that a value of 5 gave much better results for new user recommendation.
  Our key results are not sensitive to minor changes of algorithm parameters.

@@

  \paragraph{Accuracy and Ranking Metrics}
  {\bf RMSE} and {\bf nDCG} evaluate the recommender based on its prediction for each item.
  These evaluate the accuracy of the predictions and the ability of the recommender to rank the list of all items.
  We find that these two metrics perform quite similarly in our evaluation.

@@

  \paragraph{Monotonic Recommendation Metrics}
  %TODO: daniel comment: check for consistency with the user vs. users.
  These metrics evaluate the top 20 recommendations produced by the recommender.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.
  We refer to these as monotonic metrics as there is a clear good direction for the metric.

@@

  \paragraph{Non-monotonic Recommendation Metrics}
  {\bf SeenItems@20}, {\bf AveragePopularity@20}, {\bf AILS@20}, and {\bf Spread@20} are all non-monotonic measures of the user's recommendations.
  These metrics are harder to apply than the monotonic measures of recommendation quality.
  Unlike the previous metrics, we expect that users will be most satisfied by moderate values of these metrics, and that both extremes can lead to a negative user experience.

@@

  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the ItemItem recommender.

  The {\bf Spread@20} metric is designed to measure how well the recommender spreads its attention across many items.
  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
  Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
  When Spread is large we know that the recommender is recommending many items with a relatively even distribution.
  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
  In this way spread is related to item coverage metrics.
  Spread however looks at how often each item is recommended, not just whether or not it is recommended.

@@

  \paragraph{Expectations about algorithm performance}
  % provide expectations.
  We expect that the ItemBaseline algorithm will have approximately constant performance on all metrics.
  Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
  That said, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.

@@

  % describe issues with solution
  Unfortunately, we found that this strategy causes biases in our evaluation.
  In particular, this strategy led to the size of the test sets changing as a function of the simulated profile size.
  As we kept more ratings for the training set, we would have fewer ratings in the test set.
  This biased our results; some metrics such as nDCG proved to be sensitive to the size of the test set.
  As we compared different profile sizes we would see trends that were solely a property of the changing test size.

@@

\section{Results}

@@

  % Recommendation quality metrics
    % NDCG
  \paragraph{nDCG}
  Figure \ref{fig:ndcg} shows the algorithms performance on the nDCG metric.
  The results for nDCG are very similar to the RMSE results.
  Both UserUser and SVD take an initial hit to accuracy at the first rating.
  This time UserUser takes much longer to recover and only beats the baseline after around 12 ratings.
  Again ItemItem performs quite poorly.
  This leaves SVD performing the best, passing the baseline after about 3 ratings.
  Again we find that our best algorithm for the first few ratings is our baseline.

@@

  \paragraph{SeenItems@20}
  We address this non-monotonic metric out of turn here as it provides important context for the next two metrics.
  The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
  The average number of test set items in recommendations shows a similar trend to the MAP and precision metrics, with UserUser getting by far the fewest seen items, and the baseline getting the most.
  As UserUser gets less than one seen item on average we found that UserUser was not appropriate for the MeanRating@20 and RMSE@20 metrics.
  In particular, we found that the UserUser results for these metrics are highly sensitive to recommendation list size.
  Because of this issue we will not include UserUser in our next two metrics and consider the next two metrics inconclusive for UserUser.
  That said, we consider the small number of seen items to be a bad result for UserUser.
  Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?

  \paragraph{MeanRating@20}
  The second graph in figure \ref{fig:topN.rmse} shows the average rating for those items in the top 20 recommendations for each algorithm.
  The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
  SVD performs as expected, starting around the baseline for a few ratings, and then steadily improving to be the best algorithm after 3 ratings.
  Again, we find that ItemItem performs poorly until quite a few ratings are added, and does worse than baseline throughout.

@@

  \paragraph{RMSE@20}
  The third graph in figure \ref{fig:topN.rmse} shows the accuracy of our algorithms on items in the top 20 recommendations.
  Not surprisingly, this figure shows the same trends as the last one, meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
  Again, SVD performs quite well, easily beating the UserItemBaseline after only 1 rating.

@@

    % diversity plot
  \paragraph{AILS@20}
  The second graph in \ref{fig:pop} shows a similar trend to the popularity trend.
  Again, we expect that the baseline, which provides the least diverse recommendations, would be perceived as not very diverse by users.
  Quite possibly due to the preference for obscure movies, UserUser seems to produce the most diverse lists of any algorithm.
  We expect that the UserUser recommendation lists might be too diverse.
  Unlike popularity, SVD and ItemItem seem to perform equivalently in terms of diversity.
  We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing inter-list similarity) as more ratings are entered.

  % spread plot
  \paragraph{Spread@20}
  The third graph in figure \ref{fig:pop} shows the spread of the recommendations from each algorithm.
  As expected the baseline performs worst on spread, as it recommends essentially the same items to each user.
  SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
  UserUser and ItemItem start at around the same value as each other, however, after only two ratings they have crossed, as ItemItem quickly increases to have the highest spread of all the algorithms, and UserUser quickly decreases into second place.
  Interestingly, only UserUser shows a trend to have less spread as it learns more about the user, ItemItem and SVD both show the expected trend of increasing Spread as they learn more.
  This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
  More research will be needed to understand this trend.

@@

  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{4.5em}|p{4.5em}|p{7.5em}|p{4.5em}|}
      \hline
      {\bf Algorithm} & {\bf Prediction accuracy} & {\bf Recommendation quality} & {\bf Other properties} \\\hline
      ItemItem        & Poor                      & Poor                         & Good                   \\\hline
      UserUser        & Good                      & Inconclusive                 & Poor                   \\\hline
      SVD             & Good                      & Good                         & Good                   \\\hline
    \end{tabular}
    \caption{Summary of algorithm behavior for new users}
    \label{tbl:results}
  \end{table}

>>>>

  The ideal way to study this issue would be a live study conducted on people who are new to a recommender system.
  Users could be randomly assigned an algorithm and then surveyed after a random number of ratings to establish what they thought of the new user experience. 
  Unfortunately, there are many algorithms and new users are a scarce resource.

  Fortunately, there has been a growing trend in recommender system research to perform multiple metric offline evaluations of recommender systems.
  This work comes out of a pushback against early recommender evaluations which only evaluated recommenders in terms of their predictive accuracy \cite{accurateIsNotEnough}.
  The response of this pushback has been the generation of many different metrics which allow us to more carefully understand the separate aspects of a recommender's output \cite{LathiaTemporal, CremonesiTopN, zieglerDiversity}.
  %TODO: I want to cite more papers here, I don't think I have good ones either. Joe would probably know.
  This strategy allows us to dig much deeper into the differences between algorithms, looking past simple measures of accuracy to more meaningful dimensions such as how novel the recommendations are, or how frequently the recommendations change for any user.
  This allows us to capture the qualitative tradeoffs between algorithms which are otherwise quite similar, which helps us choose algorithms that are better suited for a particular domain.

@@

  Ideally we would measure new user experience with a live survey on users who are just joining a recommender service.
  Because new users are a scarce resource we opt to first explore the new user experience in an offline setting.
  This allows us to test many more profile sizes and algorithms than an online study.
  We will adapt the standard train and test methodology so that test users have only a specified number of training ratings.
  By running this approach for various profile sizes we can generate plots of performance against profile size.
  This lets us understand both how the algorithms compare on these metrics, and how the algorithm performance changes with the number of ratings provided.

@@

  Due to the low profile sizes involved in our evaluation, not every algorithm was able to make predictions for every user-item pair.
  In these cases, we use the UserItemBaseline prediction as a fallback.
  This ensured that our prediction based metrics were always evaluated over the same number of items.
  We found this unnecessary for recommendations as each algorithm was able to make a complete recommendation with only one rating.
  Therefore we form recommendations as the top 20 items with predictions from the algorithm which were not in the training set. 
  More information about these algorithms can be found in table \ref{tbl:algo}.
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

@@

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was tuned based off of Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation is on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning using Ekstrand's values as a starting point.
  We tuned based on RMSE performance.
  While tuning, we found that the previous damping parameter used for the baselines to be too large.
  We found that a value of 5 gave much better results for new user recommendation.
  Our key results are not sensitive to minor changes of algorithm parameters.

@@

  \vspace{-0.5em}
  \paragraph{Accuracy and Ranking Metrics}
  {\bf RMSE} and {\bf nDCG} evaluate the recommender based on its prediction for each item.
  These evaluate the accuracy of the predictions and the ability of the recommender to rank the list of all items.
  We find that these two metrics perform quite similarly in our evaluation.

@@

  \vspace{-0.5em}
  \paragraph{Monotonic Recommendation Metrics}
  %TODO: daniel comment: check for consistency with the user vs. users.
  These metrics evaluate the top 20 recommendations produced by the recommender.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.
  We refer to these as monotonic metrics as there is a clear good direction for the metric.

@@

  \vspace{-0.5em}
  \paragraph{Non-monotonic Recommendation Metrics}
  {\bf SeenItems@20}, {\bf AveragePopularity@20}, {\bf AILS@20}, and {\bf Spread@20} are all non-monotonic measures of the user's recommendations.
  These metrics are harder to apply than the monotonic measures of recommendation quality.
  Unlike the previous metrics, we expect that users will be most satisfied by moderate values of these metrics, and that both extremes can lead to a negative user experience.

@@

  The {\bf AILS@20} metric is based on the Intra-List Similarity (ILS) metric introduced by Ziegler et al. \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the ItemItem recommender.

  The {\bf Spread@20} metric is designed to measure how well the recommender spreads its attention across many items.
  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
  Like diversity, however, we expect that it is not possible to reach complete spread (recommending each item an equal number of times) without making avoidably bad recommendations.
  When Spread is large we know that the recommender is recommending many items with a relatively even distribution.
  When Spread is small, the recommender is focusing on a small set of items which it recommends to every user.
  In this way spread is similar to an item coverage metric, but provides a more nuanced measure of item coverage  as it considers how often each item is recommended.

@@

  \vspace{-0.5em}
  \paragraph{Expectations about algorithm performance}
  % provide expectations.
  We expect that the ItemBaseline algorithm will have approximately constant performance on all metrics.
  Since the ItemBaseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better than it.
  That said, we expect the improvement over the ItemBaseline to be gradual, with larger improvements only happening with more data.

@@

  % describe issues with solution
  Unfortunately, we found that this strategy causes biases in our evaluation.
  In particular, this strategy led to the size of the test sets changing as a function of the simulated profile size.
  As we kept more ratings for the training set, we would have fewer ratings in the test set.
  This biased our results; some metrics such as nDCG proved to be sensitive to the size of the test set.
  As we compared different profile sizes we would see trends that were solely a property of the changing test set size.

@@

\pagebreak
\section{Results}

@@

  % Recommendation quality metrics
    % NDCG
  \vspace{-0.5em}
  \paragraph{nDCG}
  Figure \ref{fig:ndcg} shows the algorithms performance on the nDCG metric.
  The results for nDCG are very similar to the RMSE results.
  Both UserUser and SVD take an initial hit to accuracy at the first rating.
  This time UserUser takes much longer to recover and only beats the baseline after around 12 ratings.
  Again ItemItem performs quite poorly.
  This leaves SVD performing the best, passing the baseline after about 3 ratings.
  Again we find that our best algorithm for the first few ratings is our baseline.

@@

  \vspace{-0.5em}
  \paragraph{SeenItems@20}
  We address this non-monotonic metric out of turn here as it provides important context for the next two metrics.
  The first graph in figure \ref{fig:topN.rmse} shows the average number of items in the top 20 recommendations for each algorithm that were also in the test set.
  The average number of test set items in recommendations shows a similar trend to the MAP and precision metrics, with UserUser getting by far the fewest seen items, and the baseline getting the most.
  As UserUser gets less than one seen item on average we found that UserUser was not appropriate for the MeanRating@20 and RMSE@20 metrics.
  In particular, we found that the UserUser results for these metrics are highly sensitive to recommendation list size.
  Because of this issue we will not include UserUser in our next two metrics and consider the next two metrics inconclusive for UserUser.
  That said, we consider the small number of seen items to be a bad result for UserUser.
  Put simply, if we cannot reliably estimate if the recommendations are reasonable, how can we expect a user to do so?

  \vspace{-0.5em}
  \paragraph{MeanRating@20}
  The second graph in figure \ref{fig:topN.rmse} shows the average rating for those items in the top 20 recommendations for each algorithm.
  The first thing to note about this graph is that all of these algorithms are doing a reasonable job of recommending items the user might like to watch, with all algorithms averaging above 4 stars (implying that the recommended items are, at least on average, enjoyable).
  SVD performs as expected, starting around the baseline for a few ratings, and then steadily improving to be the best algorithm after 3 ratings.
  Again, we find that ItemItem performs poorly until quite a few ratings are added, and does worse than baseline throughout.

@@

  \vspace{-0.5em}
  \paragraph{RMSE@20}
  The third graph in figure \ref{fig:topN.rmse} shows the accuracy of our algorithms on items in the top 20 recommendations.
  Not surprisingly, this figure shows the same trends as the last one, meaning that the algorithms whose recommendations were (on average) most likable, were those whose recommendations were most accurate.
  Again, SVD performs quite well, easily beating the UserItemBaseline after only 1 rating.

@@

    % diversity plot
  \vspace{-0.5em}
  \paragraph{AILS@20}
  The second graph in \ref{fig:pop} shows a similar trend to the popularity trend.
  Again, we expect that the baseline, which provides the least diverse recommendations, would be perceived as not very diverse by users.
  Quite possibly due to the preference for obscure movies, UserUser seems to produce the most diverse lists of any algorithm.
  We expect that the UserUser recommendation lists might be too diverse.
  Unlike popularity, SVD and ItemItem seem to perform equivalently in terms of diversity.
  We also see that all three collaborative filtering algorithms show a trend of increasing diversity (decreasing inter-list similarity) as more ratings are entered.

  % spread plot
  \vspace{-0.5em}
  \paragraph{Spread@20}
  The third graph in figure \ref{fig:pop} shows the spread of the recommendations from each algorithm.
  As expected the baseline performs worst on spread, as it recommends essentially the same items to each user.
  SVD performs mildly better than the baseline, and shows definite improvement as more ratings are added.
  UserUser and ItemItem start at around the same value as each other, however, after only two ratings they have crossed, as ItemItem quickly increases to have the highest spread of all the algorithms, and UserUser quickly decreases into second place.
  Interestingly, only UserUser shows a trend to have less spread as it learns more about the user, ItemItem and SVD both show the expected trend of increasing Spread as they learn more.
  This could be some form of a regression towards the mean, where as it learns more about users its recommendations become less personalized to the users individual quirks.
  More research will be needed to understand this trend.

@@

  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{4.5em}|p{4.5em}|p{7.5em}|p{4.5em}|}
      \hline
      { Algorithm} & { Prediction accuracy} & { Recommendation quality} & { Other properties} \\\hline
      ItemItem        & Poor                      & Poor                         & Good                   \\\hline
      UserUser        & Good                      & Inconclusive                 & Poor                   \\\hline
      SVD             & Good                      & Good                         & Good                   \\\hline
    \end{tabular}
    \caption{Summary of algorithm behavior for new users}
    \label{tbl:results}
  \end{table}
